---
title: "Analysing the sensitivity and specificity of ASVs for discriminating between genomes"
author: "Callum Thomas"
date: "2022-09-30"
output: 
  github_document:
    html_preview: true
editor_options: 
  chunk_output_type: console
---

```{r library loading, message=F, warning=F}
library(tidyverse)
library(here)
library(scales)
```
### Need to determine the number of *rrn* operons across genomes.

```{r import}
count_tibble <- read_tsv(here("data/processed/rrnDB.count_tibble"),
                         col_types = "cccd")
```

How many rrn copies are in each genome for the full-length sequences (v1-9)?

```{r n_rrn}
count_tibble %>% 
  filter(region == "v19") %>%
  group_by(genome) %>% 
  summarise(n_rrn = sum(count), .groups="drop") %>%
  ggplot(aes(x = n_rrn)) + geom_histogram(binwidth = 1)
```

What do these numbers look like as a fraction of all genomes?
```{r fraction_rrn}
count_tibble %>% 
  filter(region == "v19") %>% 
  group_by(genome) %>% 
  summarise(n_rrn = sum(count), .groups="drop") %>%
  count(n_rrn) %>% 
  mutate(fraction = n/sum(n))
```

We see that most genomes actually have more than one copy of the *rrn* operon. 
I wonder if those different copies are the same sequence/ASV?

### What is the number of ASVs per genome?

Considering that most genomes have multiple copies of the *rrn* operon, we need
to know whether they all have the same ASV. Otherwise, we run the risk of 
splitting a single genome into multiple ASVs.

```{r n_ASV, message=FALSE}
count_tibble %>% 
  group_by(region, genome) %>% 
  summarise(n_asv = n(), n_rrn = sum(count), .groups="drop") %>% 
  group_by(region, n_rrn) %>% 
  summarise(med_n_asv = median(n_asv),
            mean_n_asv = mean(n_asv),
            lq_n_asv = quantile(n_asv, prob=0.25),
            uq_n_asv = quantile(n_asv, prob = 0.75))

count_tibble %>%
  group_by(region, genome) %>% 
  summarise(n_asv = n(), n_rrn = sum(count), .groups="drop") %>% 
  ggplot(aes(x = n_rrn, y = n_asv, color = region)) + geom_smooth(method = "lm")

```

Surprisingly, the number of ASVs increases at a rate of about 2 ASVs per 3
copies of the *rrn* operon in the genome for the full-length sequence. The sub-
regions of the gene have fewer ASVs per *rrn* operon.

### How many genomes does each ASV appear in?

Are ASVs unique to genomes they are found in?

```{r specificity, message=FALSE}
(ASV_per_genome <- count_tibble %>% 
  group_by(region, asv) %>% 
  summarise(n_genome = n()) %>% 
  count(n_genome) %>% 
  mutate(fraction = n/sum(n)) %>% 
  filter(n_genome == 1))
```

We see that (for full-length sequences) `r format(ASV_per_genome$fraction[1]*100, digits =4)`
% of ASVs were unique to a genome. For the sub-regions, the percentage of ASVs
was lower: `r format(ASV_per_genome$fraction[2]*100, digits =4)`% for region 3-4,
`r format(ASV_per_genome$fraction[3]*100, digits =4)`% for region 4 and 
`r format(ASV_per_genome$fraction[4]*100, digits =4)`% for region 4-5. 

#### To be determined:
- Can we correct for overrepresentation?
- Consider analysis at species, genus, family, etc. levels?
- Consider more broad definition of an ASV
